Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-04 paper/paper/SPIE/article.tex
[WIP]

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{tocloft}

@@

@@

\textbf{NOTE}: Forget academia for a moment, and write like its a blog. Then go back and add
the citations and ajust the tone.

@@

At its most rudimentary, RMT simply describes the statistically-expected behaviour of the
eigenvalues of certain classes—or \textit{ensembles}—of random matrices under certain assumptions or
conditions. A random matrix is, generally, a matrix in which the entries can be considered to be
independently and identically distributed (iid), and where the iid distribution is a classic
mathemtical distribution. For example, one of the simplest ensembles [cite handbook] is the Gaussian
Unitary Ensemble (GUE), wherein [desc of ensemble properties] and which appears to be useful in
describing [real-world examples].

@@

\subsection{Eigenfeatures}

@@

@@

To examine the potential utility of RMT-derived features across a wide
variety of data, and not to decisively put forward or refute any specific hypotheses relating RMT to
psychological processes, we examine a wide selection of fMRI datasets available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36).  This allows for a broad assessment of the
potential of voxelwise RMT features.

However, since previous studies employing RMT for feature extraction have partially interpreted
their findings with respect to attentional processes
\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we have also elected to use
datasets that involve attention in some way. This includes measures of attention or vigilance, or
comparisons between a condition involving more attention and focus (e.g. a task state) or conditions
which should involve less attention or focus (e.g. resting-state, pain medication vs. none).

\subsection{Dataset: PSYCH}
The PSYCH Dataset  is detailed in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a
battery of psychological measurements, and various resting-state fMRI (rs-fMRI) whole-brain 7-Tesla
scans were obtained. For the purpose of keeping with the ``attention'' theme of the analyses in this
paper, three pairings of subgroups were constructed via z-score splits of the various sum-scores of
scale items involving concepts closely related to attention.

@@

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label ``REFLECT'', due to the ``reflective” nature of the
cognitive tasks) is described in full detail by \cite{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table \ref{table:2}. Tasks
involved viewing various affective images (see \cite{sprengPatternsBrainActivity2010} for details)
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as ``attending'', in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast \cite{kunduMultiechoFMRIReview2017}. In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's \(i\)th signal value
from echo \(j\) as \(t_{j,i}\), then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

This creates a physically-meaningful 4D image that contains the full, original multi-echo fMRI
time-series data without any alterations. It should be noted that the purpose of these combination
methods is not to create a 4D volume that allows optimal detection of the purely task-linked BOLD
signal, but to ensure all scan information was used with minimal additional analytic assumptions.

@@

\subsubsection{Dataset 5}
Dataset 5 (referred to in figures and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components see figure 1 of \cite{schapiroHumanHippocampalReplay2018}. After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of
the analysis of this paper. Resting state scans were included in the low-attention ``rest''
subgroup.

@@

>>>>

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}

@@

Also

@@

\textbf{NOTE}: Forget academia for a moment, and write like its a blog. Then go back and add
the citations and adjust the tone.

@@

At its most rudimentary, RMT simply describes the statistically-expected behaviour of the
eigenvalues of some classes—or \textit{ensembles}—of random matrices under certain assumptions or
conditions. A random matrix is, generally, a matrix in which the entries can be considered to be
independently and identically distributed (\textit{iid}). For example, one of the simplest ensembles [cite handbook] is the Gaussian
Unitary Ensemble (GUE), wherein [desc of ensemble properties] and which appears to be useful in
describing [real-world examples].

@@

\subsection{Motivating Eigenfeatures}

@@

In addition, the functional connectivity - the matrix of correlations of various collections of
voxels of an fMRI scan - is a priori a useful representation of the fMRI data. However, the full
correlation matrix between all voxels is itself often computationally infeasible to work with, and
thus is reduced in various ways prior to being used in analyses.

Typical reductions include the use of a ``seed'' voxel or collection of voxels (i.e. ROI): the mean
signal over that ROI is correlated with all other N ROIs of interest to generate a reduced functional
connectivity matrix (or image) with one correlation value at each non-seed ROI. This sort of reduction reduces the functional connectivity to N values. This approach strongly
biases the analysis to the seed ROI.

Likewise, one can work with ROI mean signals, and work with the \(N \times N\) matrix of
these ROI correlations. This analysis [maybe cite Poldrack book for this?] is, \textit{a priori},
sensitive to the choice of ROIs, and, in the case of anatomical atlases / parcellations, may also
lack theoretical justification.

A major disadvantage of both the above approaches, besides requiring parcellation into ROIs, is
that most such parcellation methods are atlas-based, and thus require image registration, and,
general, an entire preprocessing pipeline to ensure that registration is successful.

RMT suggests a potentially useful reduction in the form of the eigenvalues of the voxelwise
functional connectivity. An \(N \times t\) matrix \(\bf{A}\) of \(N\) time series of length \(t\),
and where \(t \ll N\) will have a symmetric correlation (or covariance) matrix with \(t - 1\)
non-zero, positive real-valued eigenvalues \(\Lambda\). These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance matrix.

@@

\subsection{Overview}

In order not to inundate readers with dataset details, and because this is an exploratory
investigation of RMT which is not committed to any specific theory, we refer the reader
to the original publications and/or data releases when greater detail is needed, and
highlight only the most basic aspects of each here.

We selected fMRI data publicly available on the \href{https://openneuro.org/}{OpenNeuro} platform
(36). Selection criteria was somewhat subjective, but we required (1), that all scans have the same
spatial and temporal resolutions, (2), the the dataset comprise 10+ human subjects, and (3), that it
was reasonable to classify at the level of an entire scan, i.e. neither task nor event timings were
needed to determine class membership. This yielded seven datasets comprising XX total subjects, and
over 5000 total fMRI volumes.

\subsubsection{Aging}
This dataset\cite{wahlheimIntrinsicFunctionalConnectivity2021} included 34 subjects with a mean age
of 22 years (range: 18-32 years), and 28 subjects with a mean age of 70 years (range: 61-80 years).

% The final sample included 34 younger adults (1832 years old, Mage = 22.21, SD = 3.65; 20 female)
% and 28 older adults (61-80 years old, Mage = 69.82, SD = 5.64; 20 female). Table 1 shows that
% younger and older adults had comparable education and working memory capacity, the latter measured
% by forward and backward digit span (48). Older adults had higher vocabulary scores (49) and slower
% processing speed (50) than younger adults.
% Resting-state data comprised 300 measurements collected over one 10-minute run. This scan duration
% produces reliable functional connectivity estimates within and across participants (57).
% Participants were instructed to remain still and awake with their eyes open. No stimuli appeared
% during this scan, and the display was black. Functional scans were collected using an echo-planar
% image sequence sensitive to blood-oxygen-level-dependent (BOLD) contrast (T2*; 32 slices with 4.0
% mm thickness and no skip, TE = 30 ms, TR = 2000 ms, flip angle = 70, FOV = 220 mm, matrix size =
% 74 × 74 × 32 voxels, A/P phase encoding direction). Slices were collected in a descending order
% that covered the entire cortex and partial cerebellum. At the beginning of the resting-state scan,
% the scanner acquired and discarded two dummy scans.

\subsubsection{Bilinguality}

This dataset\cite{ds001747:1.0.0, goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting state. The original study
grouped participants into three classes: early vs late bilinguals vs monolingual controls, and found
connectivity differences only between the early and late
bilinguals\cite{goldcarrieelizabethExploringRestingState2018}. We compared for our two classes just
"bilingual" vs "monolingual". The data is most notable for the short TR (875ms) combined with long
acquisition time (12 minutes).

% Resting state fMRI scans were performed using a T2*-weighted EPI sequence with the following
% parameters: TR=875ms; TE=43.6ms; slice thickness=1.8mm; acquisition matrix=100 × 100; flip
% angle=55°; number of slices=72; field of view=180 × 180mm; voxel size=1.8 × 1.8 × 1.8mm;
% multi-band factor=8; volumes per run=823 (total scan time=12min). EPI scans were oriented
% parallel to the long axis of the hippocampus.

\subsubsection{Learning}

This dataset\cite{schapiroHumanHippocampalReplay2018, schapiroHumanHippocampalReplay2020}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The task is complex and difficult to summarize here adequately, but involved
multiple phases where subjects were presented with various related abstract images, and then later tested for their memory of certain aspects of the learned images. For our study, we simply note that these task scans ought to be broadly more attentionally  and cognitively demanding than the resting state, and instead simply try to classify task vs. test scans using eigenfeatures.

\subsubsection{Dataset: PSYCH}
The PSYCH Dataset  is detailed in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a
battery of psychological measurements, and various resting-state fMRI (rs-fMRI) whole-brain 7-Tesla
scans were obtained. For the purpose of keeping with the ``attention'' theme of the analyses in this
paper, three pairings of subgroups were constructed via z-score splits of the various sum-scores of
scale items involving concepts closely related to attention.

@@

@@

\subsubsection{Dataset 5}

@@

\section{Results}
\label{sect:results}

First not non-predictable data, and then no longer discuss it in subsequent features.

\subsection{Normalization}
NO meaningful effect of normalization on AUROC distributions at level of gross feature, or feature
group, so can safely ignore this.
NO meaningful effect of normalization for any feature group by classifier either, so
normalization doesn't seem to matter here.
Maybe add table of correlations by groupings here

\subsection{Classifier}
When the data is predictable, timeseries features almost all have the mode / max <= 0.5 for all
classifiers and subgroups, whereas eigenfeatures (RMT, eigenvalue only, eigenvalue + RMT) more often
than not have mode greater than 0.5, and bulk of the distribution > 0.5.

KNN3 consistently has least favourable AUROC distributions across subgroups and features
